\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries

### 2015-09-02 swmpr_manu.tex
most of the formatting is done, ready to butcher text

\title{SWMPr: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
\author{by Marcus W Beck}

@@

Two basic approaches using SWMPr are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using CDMO web services. To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.  The \texttt{site\_codes} or \texttt{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.

% table for retrieval functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?all\\_params} for individual functions or \\texttt{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Retrieval functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?all\_params} for individual functions or \texttt{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \texttt{exportAllParamsXMLNew} function on web services.\tabularnewline
\texttt{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \texttt{exportAllParamsDateRangeXMLNew}.\tabularnewline
\texttt{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
\texttt{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \texttt{exportSingleParamXMLNew} function on web services.\tabularnewline
\texttt{site\_codes}&Metadata for all stations, wrapper to \texttt{exportStationCodesXMLNew} function on web services.\tabularnewline
\texttt{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \texttt{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

@@

% attributes table
%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\texttt{swmpr} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Attributes of a \texttt{swmpr} object that describe characteristics of the data.\label{tab:attributes}} 
\begin{center}
\begin{tabular}{lp{0.75in}p{3.25in}}
\hline\hline
\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{names}&character&Column names of the entire data set, inherited from the \texttt{data.frame} object class.\tabularnewline
\texttt{row.names}&integer&Row names of the data set, inherited from the \texttt{data.frame} object class.\tabularnewline
\texttt{class}&character&Class of the data object indicating \texttt{swmpr} and \texttt{data.frame}.\tabularnewline
\texttt{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
\texttt{parameters}&character&Character vector of column names for data parameters, e.g., \texttt{`do\_mgl'}, \texttt{`turb'}, etc.\tabularnewline
\texttt{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
\texttt{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
\texttt{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
\texttt{stamp\_class}&character&Class of the \texttt{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
\hline
\end{tabular}\end{center}

\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \texttt{swmpr} objects and may not be cities in the United States.  For example, ``America/Jamaica" is used for Eastern Standard Time.\end{table}

The \texttt{swmpr} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system, \cite{Wickham14}), such that specific methods for generic functions are developed for the object class.  A \texttt{swmpr} object also secondarily inherits methods from the \texttt{data.frame} class, such that common \texttt{data.frame} methods available in R also apply to \texttt{swmpr} objects.  Available methods for the \texttt{swmpr} class are described below and can also be viewed:

@@

The \texttt{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \texttt{swmpr} object will have a corresponding QAQC column of the same name with the added prefix \texttt{f\_} (e.g., \texttt{do\_mgl}, \texttt{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.  The \texttt{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \texttt{0} QAQC flag (i.e., passed initial CDMO checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \texttt{NA} (not available) values for observations that do not meet the criteria specified in the function call.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \\texttt{?comb} for individual functions or \\texttt{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Organizing functions available from the SWMPr package. Full documentation for each function is in the help file (e.g., execute \texttt{?comb} for individual functions or \texttt{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{comb}&Combines \texttt{swmpr} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.\tabularnewline
\texttt{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \texttt{swmpr} object.  Only applies if QAQC columns are present. \tabularnewline
\texttt{qaqcchk}&View a summary of the number of observations in a \texttt{swmpr} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.\tabularnewline
\texttt{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\texttt{setstep}&Format data from a \texttt{swmpr} object to a continuous time series at a given timestep.  The function is also used in \texttt{comb}.\tabularnewline
\texttt{subset}&Subset by dates and/or columns for a \texttt{swmpr} object.  This is a method passed to the generic \texttt{subset} function provided in the base package.\tabularnewline
\hline
\end{tabular}\end{center}

@@

Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \texttt{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call, including treatment of \texttt{NA} values, should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.

@@

A subset method added to the existing generic \texttt{subset} function in R is available for \texttt{swmpr} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the \texttt{operator} argument, such as \texttt{`>'} or \texttt{`<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{`2012-11-31 12:00'}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The \texttt{setstep} function formats a \texttt{swmpr} object to a continuous time series at a given time step.  This function is usually not necessary because most stations collect data at 15 minute intervals, but it can be used to combine data or convert an existing time series to a different interval.  The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.  Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.  Likewise, the default value for \texttt{differ} is one half the time step.  Time steps that do not match any existing data within the limits of the \texttt{differ} argument are not discarded, although a corresponding data value will not be assigned.

@@

The \texttt{comb} function is used to combine multiple \texttt{swmpr} objects into a single object with a continuous time series at a given step.  The \texttt{setstep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.  All arguments must be called explicitly since an arbitrary number of \texttt{swmpr} objects can be used as input to \texttt{comb}.  The function combines data by creating a master time series that is used to iteratively merge all \texttt{swmpr} objects.  The time series for merging depends on the value passed to the \texttt{method} argument.  Passing \texttt{`union'} to \texttt{method} will create a time series that is continuous from the earliest and latest dates for all input objects.  Passing \texttt{`intersect'} to \texttt{method} will create a continuous time series from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to \texttt{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.  Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (Table~\ref{tab:analyze}).  The general purpose tools are \texttt{swmpr} methods that were developed for existing generic functions in the R base installation or relevant packages (SWMPr imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}). These functions include \texttt{swmpr} methods for \texttt{aggreswmp}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \texttt{plot}, \texttt{lines}, and \texttt{hist} methods.  Decomposition functions, \texttt{decomp} and \texttt{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \texttt{ecometab} and \texttt{plot\_metab} functions.  The analysis functions may or may not return a \texttt{swmpr} object depending on whether further processing with \texttt{swmpr} methods is possible from the output.

The \texttt{aggreswmp} function aggregates parameter data for a \texttt{swmpr} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.  A \texttt{swmpr} object is returned for the aggregated data, although the \texttt{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects created in the data.table package \cite{Dowle14}.  Additionally, the method of treating \texttt{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \\texttt{?aggreswmp} for individual functions or \\texttt{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the SWMPr package.  Full documentation for each function is in the help file (e.g., execute \texttt{?aggreswmp} for individual functions or \texttt{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\texttt{aggreswmp}&Aggregate \texttt{swmpr} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\texttt{aggremetab}&Aggregate metabolism data from a \texttt{swmpr} object.  This is primarly used within \texttt{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\texttt{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
\texttt{decomp}&Decompose a \texttt{swmpr} time series into trend, seasonal, and residual components.  This is a simple wrapper to \texttt{decompose} \cite{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\texttt{decomp\_cj}&Decompose a \texttt{swmpr} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \texttt{decompTs} in the wq package \cite{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\texttt{hist}&Plot a histogram for a \texttt{swmpr} object.\tabularnewline
\texttt{lines}&Add lines to an existing plot created with \texttt{plot}.\tabularnewline
\texttt{map\_reserve}&Create a map of all stations in a reserve using the ggmap package \cite{Kahle13}.\tabularnewline
\texttt{na.approx}&Linearly interpolate missing data (\texttt{NA} values) in a \texttt{swmpr} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
\texttt{plot}&Plot a univariate  time series for a \texttt{swmpr} object.  The parameter name must be specified.\tabularnewline
\texttt{plot\_metab}&Plot ecosystem metabolism estimates after running \texttt{ecometab} on a combined \texttt{swmpr} object.\tabularnewline
\texttt{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
\texttt{smoother}&Smooth \texttt{swmpr} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \texttt{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

Time series can be smoothed to better characterize a signal from noisy data (Figure~\ref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a \texttt{swmpr} object using a specified window size.  The \texttt{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.  The \texttt{sides} argument specifies how the average is calculated for each observation.  Setting \texttt{sides = 1} will filter observations within the window that are previous to the current observation, whereas \texttt{sides = 2} will filter observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.

@@

{\centering \includegraphics[width=0.00\textwidth]{figure/smooth_ex-1}

@@

\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \texttt{smoother} function.}\label{fig:smooth_ex}
\end{figure}

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach implemented in the SWMPr package is linear interpolation using the \texttt{na.approx} function (Figure~\ref{fig:interp_ex}).  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.  An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.  The \texttt{maxgap} argument passed to \texttt{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.

@@

{\centering \includegraphics[width=0.00\textwidth]{figure/interp_ex-1}

@@

\caption[Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \texttt{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.}\label{fig:interp_ex}
\end{figure}

>>>>

\title{\pkg{SWMPr}: An R Package for Retrieving, Organizing, and Analyzing Environmental Data for Estuaries}
\author{by Marcus W Beck}

@@

Two basic approaches using \pkg{SWMPr} are available to import SWMP data into R, either through direct download or by importing local data (Table~\ref{tab:retrieve}). First, functions from the package can be used to import the data directly from the online server using CDMO web services. To do so, the IP address for the computer making the request must be registered by following instructions on the CDMO website.  The \code{site\_codes} or \code{site\_codes\_ind} functions can be used to view the available metadata after a computer is registered with CDMO.

% table for retrieval functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Retrieval functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?all\\_params} for individual functions or \\code{help.search(`retrieve', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:retrieve",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Retrieval functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?all\_params} for individual functions or \code{help.search(`retrieve', package = `SWMPr')} for all).\label{tab:retrieve}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{all\_params}&Retrieve records starting with the most recent at a given station, all parameters.  Wrapper to \code{exportAllParamsXMLNew} function on web services.\tabularnewline
\code{all\_params\_dtrng}&Retrieve records of all parameters within a given date range for a station.  Optional argument for a single parameter. Wrapper to \code{exportAllParamsDateRangeXMLNew}.\tabularnewline
\code{import\_local}&Import files from a local path.  The files must be in a specific format, such as those returned from the CDMO using the zip downloads option.\tabularnewline
\code{single\_param}&Retrieve records for a single parameter starting with the most recent at a given station.  Wrapper to \code{exportSingleParamXMLNew} function on web services.\tabularnewline
\code{site\_codes}&Metadata for all stations, wrapper to \code{exportStationCodesXMLNew} function on web services.\tabularnewline
\code{site\_codes\_ind}&Metadata for all stations at a single site, wrapper  to \code{NERRFilterStationCodesXMLNew} function on web services.\tabularnewline
\hline
\end{tabular}\end{center}

@@

% attributes table
%latex.default(to_tab[, -1, drop = F], file = "", caption = "Attributes of a \\code{\"swmpr\"} object that describe characteristics of the data.",     rowlabel = "Attributes", colheads = c("Class", "Description"),     rowname = to_tab$Attributes, caption.loc = "top", insert.bottom = foot,     col.just = c("p{0.75in}", "p{3.25in}"), label = "tab:attributes",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Attributes of a \code{"swmpr"} object that describe characteristics of the data.\label{tab:attributes}} 
\begin{center}
\begin{tabular}{lp{0.75in}p{3.25in}}
\hline\hline
\multicolumn{1}{l}{Attributes}&\multicolumn{1}{c}{Class}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{names}&character&Column names of the entire data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{row.names}&integer&Row names of the data set, inherited from the \code{data.frame} object class.\tabularnewline
\code{class}&character&Class of the data object indicating \code{"swmpr"} and \code{data.frame}.\tabularnewline
\code{station}&character&Station identifier used by NERRS as a string with 7 or 8 characters.\tabularnewline
\code{parameters}&character&Character vector of column names for data parameters, e.g., \code{`do\_mgl'}, \code{`turb'}, etc.\tabularnewline
\code{qaqc\_cols}&logical&Indicates if QAQC columns are present in the raw data.\tabularnewline
\code{date\_rng}&POSIXct&Start and end dates for the data.\tabularnewline
\code{timezone}&character&Timezone of the station using the city/country format\textsuperscript{a}.\tabularnewline
\code{stamp\_class}&character&Class of the \code{datetimestamp} column, usually POSIXct unless data have been aggregated.\tabularnewline
\hline
\end{tabular}\end{center}

\textsuperscript{a}\footnotesize Time zones that do not observe daylight savings are used for \code{"swmpr"} objects and may not be cities in the United States.  For example, ``America/Jamaica" is used for Eastern Standard Time.\end{table}

The \code{"swmpr"} object class was created for use with the organizing and analyzing functions.  This object-oriented approach is standard for R (i.e., the S3 object  system), such that specific methods for generic functions are developed for the object class.  A \code{"swmpr"} object also secondarily inherits methods from the \code{data.frame} class, such that common \code{data.frame} methods available in R also apply to \code{"swmpr"} objects.  Available methods for the \code{"swmpr"} class are described below and can also be viewed:

@@

The \code{qaqc} function is a simple screen to retain observations from the data with specified QAQC flags (see \url{http://cdmo.baruch.sc.edu/data/qaqc.cfm}). Each parameter in the imported \code{"swmpr"} object will have a corresponding QAQC column of the same name with the added prefix \code{f\_} (e.g., \code{do\_mgl}, \code{f\_do\_mgl}).  Values in the QAQC column range from -5 to 5 to indicate the QAQC flag that was assigned by CDMO during initial processing.  The \code{qaqc} function is used to remove observations in the raw data with given flags, with the default option to retain only values with the \code{0} QAQC flag (i.e., passed initial CDMO checks).     Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. The function returns the original data with the QAQC columns removed and \code{NA} (not available) values for observations that do not meet the criteria specified in the function call.

% table for organize functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Organizing functions available from the \\pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \\code{?comb} for individual functions or \\code{help.search(`organize', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:organize",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Organizing functions available from the \pkg{SWMPr} package. Full documentation for each function is in the help file (e.g., execute \code{?comb} for individual functions or \code{help.search(`organize', package = `SWMPr')} for all).\label{tab:organize}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{comb}&Combines \code{"swmpr"} objects to a common time series using setstep, such as combining the weather, nutrients, and water quality data for a single station. Only different data types can be combined.\tabularnewline
\code{qaqc}&Remove QAQC columns and remove data based on QAQC flag values for a \code{"swmpr"} object.  Only applies if QAQC columns are present. \tabularnewline
\code{qaqcchk}&View a summary of the number of observations in a \code{"swmpr"} object that are assigned to each QAQC flag used by CDMO.  The output can be used to inform further processing.\tabularnewline
\code{rem\_reps}&Remove replicate nutrient data that occur on the same day.  The default is to average replicates.\tabularnewline
\code{setstep}&Format data from a \code{"swmpr"} object to a continuous time series at a given timestep.  The function is also used in \code{comb}.\tabularnewline
\code{subset}&Subset by dates and/or columns for a \code{"swmpr"} object.  This is a method passed to the generic \code{subset} function provided in the base installation.\tabularnewline
\hline
\end{tabular}\end{center}

@@

Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.  The \code{rem\_reps} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The \code{datetimestamp} column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defaults to the mean.  Other suggested functions include the \code{median}, \code{min}, or \code{max}.  The entire function call, including treatment of \code{NA} values, should be passed to the \code{FUN} argument (see the examples).  The function is meant to be used after \code{qaqc} processing, although it works with a warning if QAQC columns are present.

@@

A subset method added to the existing generic \code{subset} function in R is available for \code{"swmpr"} objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator as a character string passed to the \code{operator} argument, such as \code{`>'} or \code{`<='}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (e.g., `2007-01-01 06:30').  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \code{`2012-11-31 12:00'}.  Finally, the function can be used to remove rows and columns that do not contain data.

@@

The \code{setstep} function formats a \code{"swmpr"} object to a continuous time series at a given time step.  This function is usually not necessary because most stations collect data at 15 minute intervals, but it can be used to combine data or convert an existing time series to a different interval.  The first argument of the function, \code{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, \code{differ}, specifies the allowable tolerance in minutes for matching existing observations to the defined time steps in cases where the two are dissimilar.  Values for \code{differ} that are greater than one half the value of \code{timestep} are not allowed to prevent duplication of existing data.  Likewise, the default value for \code{differ} is one half the time step.  Time steps that do not match any existing data within the limits of the \code{differ} argument are not discarded, although a corresponding data value will not be assigned.

@@

The \code{comb} function is used to combine multiple \code{"swmpr"} objects into a single object with a continuous time series at a given step.  The \code{setstep} function is used internally such that \code{timestep} and \code{differ} are accepted arguments for \code{comb}.  All arguments must be called explicitly since an arbitrary number of \code{"swmpr"} objects can be used as input to \code{comb}.  The function combines data by creating a master time series that is used to iteratively merge all \code{"swmpr"} objects.  The time series for merging depends on the value passed to the \code{method} argument.  Passing \code{`union'} to \code{method} will create a time series that is continuous from the earliest and latest dates for all input objects.  Passing \code{`intersect'} to \code{method} will create a continuous time series from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to \code{method} will merge all data based on a continuous time series for the specified station, which must be present in the input data.  Currently, combining identical data types from different stations is not possible (e.g., two water quality stations from the same reserve).

@@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries (Table~\ref{tab:analyze}).  The general purpose tools are \code{"swmpr"} methods that were developed for existing generic functions in the R base installation or relevant packages (\pkg{SWMPr} imports and dependencies are listed on \href{http://cran.r-project.org/web/packages/SWMPr/index.html}{CRAN}). These functions include \code{"swmpr"} methods for \code{aggreswmp}, \code{filter}, and \code{approx} to deal with missing or noisy data and more general functions for exploratory data analysis, such as \code{plot}, \code{lines}, and \code{hist} methods.  Decomposition functions, \code{decomp} and \code{decomp\_cj}, are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. Functions to estimate and plot ecosystem metabolism from combined water quality and weather data are provided by the \code{ecometab} and \code{plot\_metab} functions.  The analysis functions may or may not return a \code{"swmpr"} object depending on whether further processing with \code{"swmpr"} methods is possible from the output.

The \code{aggreswmp} function aggregates parameter data for a \code{"swmpr"} object by set units of time.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours by a user-defined function, which defaults to the mean.  A \code{"swmpr"} object is returned for the aggregated data, although the \code{datetimestamp} vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \code{round} method for IDate objects created in the \CRANpkg{data.table} package \citep{Dowle14}.  Additionally, the method of treating \code{NA} values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned, particularly for nutrient data (see the example below).

% table for analysis functions
%latex.default(to_tab[, "Description", drop = F], file = "", caption = "Analysis functions available from the \\pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \\code{?aggreswmp} for individual functions or \\code{help.search(`analyze', package = `SWMPr')} for all).",     rowlabel = "Function", colheads = "Description", rowname = to_tab$Functions,     caption.loc = "top", col.just = c("p{3.5in}"), label = "tab:analyze",     table.env = FALSE, where = "!tbp")%
\begin{table}[!tbp]
\caption{Analysis functions available from the \pkg{SWMPr} package.  Full documentation for each function is in the help file (e.g., execute \code{?aggreswmp} for individual functions or \code{help.search(`analyze', package = `SWMPr')} for all).\label{tab:analyze}} 
\begin{center}
\begin{tabular}{lp{3.5in}}
\hline\hline
\multicolumn{1}{l}{Function}&\multicolumn{1}{c}{Description}\tabularnewline
\hline
\code{aggreswmp}&Aggregate \code{"swmpr"} objects for different time periods - years, quarters, months,  weeks, days, or hours.  The aggregation function defaults to the mean.\tabularnewline
\code{aggremetab}&Aggregate metabolism data from a \code{"swmpr"} object.  This is primarly used within \code{plot\_metab} but may be useful for simple summaries of daily metabolism data.\tabularnewline
\code{ecometab}&Estimate ecosystem metabolism for a combined water quality and weather dataset using the open-water method.\tabularnewline
\code{decomp}&Decompose a \code{"swmpr"} time series into trend, seasonal, and residual components.  This is a simple wrapper to \code{decompose} \citep{Kendall83}.  Decomposition of monthly or daily trends is possible.\tabularnewline
\code{decomp\_cj}&Decompose a \code{"swmpr"} time series into grandmean, annual, seasonal, and events components.  This is a simple wrapper to \code{decompTs} in the \CRANpkg{wq} package \citep{Jassby14}.  Only monthly decomposition is possible.\tabularnewline
\code{hist}&Plot a histogram for a \code{"swmpr"} object.\tabularnewline
\code{lines}&Add lines to an existing plot created with \code{plot}.\tabularnewline
\code{map\_reserve}&Create a map of all stations in a reserve using the \CRANpkg{ggmap} package \citep{Kahle13}.\tabularnewline
\code{na.approx}&Linearly interpolate missing data (\code{NA} values) in a \code{"swmpr"} object. The maximum gap size that is interpolated is defined by the arguments.\tabularnewline
\code{plot}&Plot a univariate  time series for a \code{"swmpr"} object.  The parameter name must be specified.\tabularnewline
\code{plot\_metab}&Plot ecosystem metabolism estimates after running \code{ecometab} on a combined \code{"swmpr"} object.\tabularnewline
\code{plot\_summary}&Create summary plots of seasonal/annual trends and anomalies for a water a single paramter of interest.\tabularnewline
\code{smoother}&Smooth \code{"swmpr"} objects with a moving window average.  Window size and sides (e.g., centered) can be specified, passed to \code{filter}.\tabularnewline
\hline
\end{tabular}\end{center}

@@

Time series can be smoothed to better characterize a signal from noisy data (Figure~\ref{fig:smooth_ex}).  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \code{smoother} function can be used to smooth parameters in a \code{"swmpr"} object using a specified window size.  The \code{window} argument specifies the number of observations included in the moving average where larger windows result in greater smoothing.  The \code{sides} argument specifies how the average is calculated for each observation.  Setting \code{sides = 1} will filter observations within the window that are previous to the current observation, whereas \code{sides = 2} will filter observations within the window centered at zero lag from the current observation. As before, the \code{params} argument specifies which parameters to smooth.

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/smooth_ex-1}

@@

\caption[Raw and smoothed dissolved oxygen data for a two-week period after using the \code{smoother} function]{Raw and smoothed dissolved oxygen data for a two-week period after using the \code{smoother} function.}\label{fig:smooth_ex}
\end{figure}

@@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach implemented in the \pkg{SWMPr} package is linear interpolation using the \code{na.approx} function (Figure~\ref{fig:interp_ex}).  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  However, the ability of the interpolated data to approximate actual trends is related to the maximum gap size between observations with missing data.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are often more accurate.  An upper limit on the maximum gap size to interpolate trends depends on the characteristics of the dataset such that a trial and error approach is appropriate for most applications.  The \code{maxgap} argument passed to \code{na.approx} defines the maximum gap size for interpolation and the following illustrates use of different maximum values to fill missing data.

@@

{\centering \includegraphics[width=0.8\textwidth]{figure/interp_ex-1}

@@

\caption[Examples illustrating use of the \code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.]{Examples illustrating use of the \code{na.approx} function to fill gaps of different sizes in a dissolved oxygen time series for a four day period.}\label{fig:interp_ex}
\end{figure}
